A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2016-03-08 Papers/InProgress/NaturalDeduction/GodProof-ND.tex

It is likely that a reduction from {\bf S5} to {\bf KB} is possible for these variants as well, since the technique we used to replace {\bf S5}'s iteration principle by Brouwer's reduction principle depends only on the structure of the final steps of the proof of Theorem 3, but not on details of the axioms.

>>>>

A reduction from {\bf S5} to {\bf KB} is possible for these variants as well, since the technique we used to replace {\bf S5}'s iteration principle by Brouwer's reduction principle depends only on the structure of the final steps of the proof of Theorem 3, but not on details of the axioms.

\section{A natural deduction proof based on Anderson's axiomatization}

%Anderson modification of the axioms in A1 equivalence removed and definition of essence altered. 
%Anderson's implication instead of equivalence. 
\begin{axiom}
\label{A1}
Either a property or its negation is positive, but not both:
$$
\all \varphi. [P(\neg \varphi) \imp \neg P(\varphi)]
$$
\end{axiom}

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(\varphi) \imp \pos \ex x.\varphi(x)]
$$
\end{theorem}

The proof is otherwise identical to the proof of theorem \ref{T1} in section \ref{sec:newproof} but the equivalence of axiom \ref{A1} does not need to be reduced to an implication by implication elimination.

\begin{definition}[Anderson's alternative Godlike]
\label{D1}
%A \emph{God-like} being possesses all positive properties:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \biimp \nec \varphi(x)]
$$
\end{definition}

\begin{corollary}
\label{C1}
Possibly, a God-like being exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}

\begin{definition}[Anderson's alternative essence with Scott's additional conjunct]
\label{D2}
%sAn \emph{essence} of an individual is a property possessed by it and necessarily implying any of its properties:
$$
\ess{\varphi}{x} \biimp \varphi(x) \wedge \all \psi. (\nec \psi(x) \biimp \nec \all x. (\varphi(x) \imp \psi(x)))
$$
\end{definition}

%At this point, G\"odel and Scott proceed to prove Theorem 2, which is then used to prove Lemma 1. However, a close inspection of Scott's proof of Lemma 1 reveals that Theorem 2 is an unnecessary detour. Therefore, we do not include it in the new proof.

\begin{definition}[Anderson's alternative where G\"odel's definition of essence is replaced by def. \ref{D2}]
\label{D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{proof} \hfill
\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$}\RightLabel{$\ex_E$}
\UIC{$[\subproof_6] \quad G(\gamma)$}
\end{prooftree}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  \dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(\gamma)]$}\RightLabel{$\forall_E$}
\UIC{$P(\lambda x.\rho(x)) \biimp \nec \rho(\gamma)$}
\AXC{$ $} \RightLabel{5}
\UIC{$\nec \rho(\gamma)$} \RightLabel{$\imp_E$}
\BIC{$P(\lambda x.\rho(x))$} 
\AXC{$ $} \RightLabel{6}
\UIC{$G(y) $}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi. [P(\lambda x.\varphi(x)) \biimp \nec \varphi(y)]$}
\UIC{$[P(\lambda x.\rho(x)) \biimp \nec \rho(y)]$}
\BIC{$\nec \rho(y)$}\RightLabel{$\nec_E$}
\UIC{$\omega':  \rho(y)$}\RightLabel{$\imp_I^6$}
\UIC{$\omega':  G(y)\imp \rho(y)$}\RightLabel{$\forall_I$}
\UIC{$\omega':  \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\nec_I$}
\UIC{$\nec \forall y. (G(y)\imp \rho(y))$}\RightLabel{$\imp_I^5$}
\UIC{$\nec \rho(\gamma)\imp \nec \forall y. (G(y)\imp \rho(y))$}
\end{prooftree}
\end{scriptsize}

\begin{scriptsize}
\begin{prooftree}
\AXC{$\nec P(\rho)\imp \nec \forall y .(G(y)\imp \rho(y))$} \RightLabel{$\imp_E$}
\AXC{$ $} \RightLabel{6}
\UIC{$\nec \rho(\gamma)$} \RightLabel{}
\UIC{$\rho(\gamma)$} \RightLabel{$\nec_E$}
\AXC{$\subproof_7$} \dashedLine
\UIC{$\omega': \rho(\gamma)\imp\Box P(\rho)$}\RightLabel{$\imp_E$}
\BIC{$\nec P(\rho)$} \RightLabel{$\imp_E$}
\BIC{$\nec \forall y .(G(y)\imp \rho(y))$}\RightLabel{$\imp_I^6$}
\UIC{$\rho(\gamma) \imp \nec \forall y.(G(y) \imp \rho(y))$}\RightLabel{$\forall_I$}
\UIC{$\forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}  
\BIC{$G(\gamma) \wedge \forall \psi.(\psi(\gamma) \imp \nec \forall y.(G(y) \imp \psi(y)))$}\dottedLine \RightLabel{D\ref{D2}}
\UIC{$[\subproof_8] \qquad \ess{G}{\gamma}$}
\end{prooftree}
\end{scriptsize}

%implication for equivalence in essence: 
\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{7}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\nec_E$}
\UIC{$\omega'':   \forall y. [G(y)\imp \varphi(y)]$}\RightLabel{$\forall_E$}
\UIC{$\omega'':   G(\gamma)\imp \varphi(\gamma)$}
\AXC{$\subproof_6$} \dashedLine
\UIC{$\omega'':  G(\gamma)$}  \RightLabel{$\imp_E$}
\BIC{$\omega'':   \varphi(\gamma)$}\RightLabel{$\nec_I$}
\UIC{$\nec \varphi(\gamma)$}\RightLabel{$\imp_I^7$}
\UIC{$\nec  \forall y. [G(y)\imp \varphi(y)]\imp \nec \varphi(\gamma)$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$\subproof_8$} \dashedLine
\UIC{$\nec \varphi(\gamma)\biimp \nec  \forall y. [G(y)\imp \varphi(y)] $}
\AXC{$ $} \RightLabel{6}
\UIC{$G(x)$}\RightLabel{$\wedge_I$}
\BIC{$\ess{G}{\gamma}$}
\AXC{Axiom \ref{A5}} \dashedLine
\UIC{$P(E)$}\dottedLine 
\AXC{$\subproof_6$} \dashedLine
\UIC{$G(\gamma)$}\dottedLine \RightLabel{D\ref{D1}}
\UIC{$\forall \varphi.(P(\varphi) \biimp \nec \varphi(\gamma))$}\RightLabel{$\forall_E$}
\UIC{$P(E) \biimp \nec E(\gamma)$}\RightLabel{$\imp_E$}
\BIC{$\nec E(\gamma)$}\RightLabel{$\nec_E$ (reflexive frame)}
\UIC{$ E(\gamma)$} \dottedLine \RightLabel{D\ref{D3}}
\UIC{$ \all \varphi.[\ess{\varphi}{\gamma} \imp \nec \ex x.\varphi(x)] $}\RightLabel{$\all_E$}
\UIC{$ \ess{G}{\gamma} \imp \nec \ex x. G(x) $}\RightLabel{$\imp_E$}
\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{small}
\end{proof}

\subsection{Why is the proof of modal collapse prevented?}

The alteration of the definition of essence in Anderson's proof prevents the standard proof of modal collapse \ref{Collapse} 
$$\vdash A \imp \nec A
$$ 
by the additional necessitation, which converts the proof of modal collapse into a proof of a tautology.
$$
\vdash \nec A \imp \nec A
$$

%For all propositions $A$, the following \emph{modal collapse} proposition is not provable with Anderson's alternative definitions with the same proof as G\"odel or Scott's:

\begin{small}
\begin{prooftree}
\AXC{Theorem 2}\dashedLine
\UIC{$\all y.[G(y) \imp \ess{G}{y}]$} \RightLabel{D2 }\doubleLine
\UIC{$\all y.[G(y) \imp G(y) \wedge \all \psi. (\nec \psi(y) \biimp \nec \all x. (G(x) \imp \psi(x)))]$}
\UIC{$\all y.[G(y) \imp \all \psi. (\nec \psi(y) \biimp \nec \all x. (G(x) \imp \psi(x)))]$} \doubleLine
\UIC{$\all y.[G(y) \imp (\nec A(y) \biimp \nec \all x. (G(x) \imp A(x)))]$} \RightLabel{$A$ is constant}
\UIC{$\all y.[G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A))]$} \doubleLine
\UIC{$[\subproof_9] \qquad \ex y.G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A))$}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ \subproof_9 $}\dashedLine
\UIC{$\ex y.G(y) \imp (\nec A \biimp \nec \all x. (G(x) \imp A))$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\ex y.G(y)$} \RightLabel{$\imp_E$}
\BIC{$ \nec A \biimp \nec \all x. (G(x) \imp A)$} \doubleLine
\UIC{$ \nec A \biimp \nec (\ex x. G(x) \imp A)$}
\AXC{$ $} \RightLabel{1}
\UIC{$\nec A$} \RightLabel{$\imp_E$}
\BIC{$ \nec (\ex x. G(x) \imp A)$} \RightLabel{$\nec_E$}
\UIC{$ \omega:  \ex x. G(x) \imp A$} \RightLabel{$\imp_E$}
\AXC{Theorem 3}\dashedLine
\UIC{$\nec \ex y.G(y)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \ex x.G(x)$} \RightLabel{$\imp_E$}
\BIC{$\omega: A$} \RightLabel{$\nec_I$}
\UIC{$ \nec A$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec A$}
\end{prooftree}
\end{small}

The exchange of the implication in the G\"odel-Scott definition of essence for an equivalence and the additional  modal necessitation of the property $A$ prevents the implication of a necessary fact from its contingent counterpart as is seen in the second part of the proof above because the hypothetical assumption of the contingent property $A$ has to be exchanged for its necessitation for the proof structure to be preserved.

\section{A natural deduction proof based on Bj\o rdal's axiomatization}

In Bj\o rdal's axiomatization godlikeness is taken as primitive, $G_B$.

\begin{definition}
\label{B:D1}
A property is positive iff it is necessarily possessed by every God-like being.
$$
P_B(\varphi) \equiv \nec \all x. [G_B(x) \imp \varphi(x)]
$$
\end{definition}

\begin{definition}
\label{B:D2}
A maximal composite of an individual's positive properties is a positive property possessed by the individual and necessarily implying every positive property possessed by the individual.
$$MCP(\varphi, x) \equiv (\varphi(x) \wedge P_B (\varphi)) \wedge \all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(\varphi(y) \imp \psi(y)))
$$
\end{definition}

\begin{definition}
\label{B:D3}
Necessary existence of an individual is the necessary exemplification of all its maximal composites.
$$
NE_B (x) \equiv \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))
$$
\end{definition}

\begin{axiom}
\label{A:A1'} If a property is positive, its negation is not positive: 
$$
\all \varphi. [P_B (\varphi) \imp \neg P_B (\neg\varphi)]
$$
\end{axiom}

\begin{axiom}
\label{A:A5'} Necessary existence is a positive property. 
$$
P_B (NE_B )
$$
\end{axiom}

The proof of the possible instantiation of $G_B$ is adjusted by proving Theorem 1 of section \ref{sec:newproof} directly for $G_B$ and not for properties in general. Corollary 1 of section \ref{sec:newproof} follows by proving the positivity of godlikeness as in lemma \ref{B:L1} below.

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
\all \varphi. [P(G_B) \imp \pos \ex x.G_B(x)]
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec \neg \ex x.G_B(x) $} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x.G_B(x) $} 
\AXC{$ $} \RightLabel{$1$}
\UIC{$\omega: G_B(x)$} \RightLabel{$\ex_I$}
\UIC{$\omega: \ex x.G_B(x) $} \RightLabel{$\imp_E$}
\BIC{$\omega:  \bot $}\RightLabel{$\imp_I^1$}
\UIC{$\omega:  \neg G_B(x) $} \RightLabel{$\imp_I$}
\UIC{$\omega:  G_B(x) \imp \neg G_B(x) $} \RightLabel{$\all_I$}
\UIC{$\omega:  \all x.(G_B(x) \imp \neg G_B(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(G_B(x) \imp \neg G_B(x)) $}\RightLabel{$\wedge_I$}
\UIC{$P(\neg G_B)$}
\AXC{Axiom 6***} \dashedLine
\UIC{$\all \varphi. [P(\neg G_B) \imp \neg P(G_B)]$} \RightLabel{$\all_E$}
\UIC{$ P( \neg G_B(x)) \imp \neg P(G_B) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(G_B) $} 
\AXC{$ $} \RightLabel{3} 
\UIC{$P(G_B)$} 
\BIC{$\bot$} \RightLabel{$\imp_I^2$}
\UIC{$ \neg \nec \neg \ex x.G_B(x)$} \doubleLine %\RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.G_B(x)$}  \RightLabel{$\imp_I^3$}
\UIC{$ P(\rho) \imp \pos \ex x.G_B(x) $} \RightLabel{$\all_I$}
\UIC{$ \all \varphi. [P(G_B) \imp \pos \ex x.G_B(x)] $}

\begin{lemma}
\label{B:L1}
If there is a God-like being, then there is a God-like being necessarily:
$$
\ex z. G_B(z) \imp \nec \ex x. G_B(x)
$$
\end{lemma}

\begin{small}
\begin{prooftree}
\AXC{$ $}
\UIC{$G_B(x) $} 
\AXC{$ $}\doubleLine
\UIC{$\nec \all x. [G_B(x) \imp G_B(x)] $}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$P_B(G_B) $}\RightLabel{ $\wedge_I$}
\BIC{$G_B(x) \wedge P_B(G_B) $}
\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{$ $}\dashedLine
\UIC{$G_B(x) \wedge P_B(G_B) $}
\AXC{$ $}\RightLabel{ $1$}
\UIC{$(\psi(x) \wedge P_B (\psi))$} \RightLabel{ $\wedge_E$}
\UIC{$P_B (\psi)$}\RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec\all y.(G_B(y) \imp \psi(y))$} \RightLabel{ $\imp_I^1$}
\UIC{$(\psi(x) \wedge P_B (\psi))\imp \nec\all y.(G_B(y) \imp \psi(y))$}\RightLabel{ $\forall_I$}
\UIC{$\all\psi.((\psi(x) \wedge P_B (\psi)) \imp \nec\all y.(G_B(y) \imp \psi(y)))$} \RightLabel{ $\wedge_I$}
\BIC{$MCP(G_B, x) $}

\end{prooftree}
\end{small}

\begin{small}
\begin{prooftree}
\AXC{Axiom \ref{A:A5'}}\dashedLine
\UIC{$P_B (NE_B)$} \RightLabel{ D\ref{B:D1}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp NE_B(x)]$}\RightLabel{ D\ref{B:D3}}\dottedLine
\UIC{$\nec \all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\nec_E$}
\UIC{$\all x. [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}  \RightLabel{ $\forall_E$}
\UIC{$ [G_B(x) \imp \all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$}
\AXC{$ $} \RightLabel{ $2$}
\UIC{$G_B(x) $}  \RightLabel{ $\imp_E$}
\BIC{$\all\varphi. (MCP(\varphi, x) \imp \nec \ex y.\varphi(y))]$} \RightLabel{ $\forall_E$}
\UIC{$(MCP(G_B, x) \imp \nec \ex y.G_B(y))]$}
\AXC{$ $}\dashedLine
\UIC{$MCP(G_B, x) $} \RightLabel{ $\imp_E$}
\BIC{$\nec \ex y.G_B(y)$}\RightLabel{ $\imp_I^2$}
\UIC{$G_B(x) \imp \nec \ex y.G_B(y)$}
\end{prooftree}
\end{small}

%ToDo: acknowledgements
